Fourier-domain modulations and delays of gravitational-wave signals

### 2017-11-03 docs/paper_method.tex
Edits to the paper - redid the plots for the response.

\usepackage{showlabels}

@@

It was suggested by a number of authors~\cite{} \SM{[C]} that the effect of the precession can be modeled, to a good approximation, by a time-dependent rotation of a non-precessing waveform. In this approach, one introduces a precessing frame following the evolution of the plane of the orbit, and approximates the waveform in this precessing frame by using a non-precessing model. Different prescriptions have been proposed for the construction of the precessing frame itself~\cite{} \SM{[C]}.

@@

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (at the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these Newtonian estimates will give useful orders of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $m=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/m^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}\label{eq:omegaphiN}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (Gm)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{G m \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
As for the leading-order time-domain amplitude of the $22$ mode, we have~\cite{} \SM{[C]}
\be\label{eq:a22N}
	a_{22}^{\rm N} (t) = \frac{2 G m \nu x}{D c^{2}} \sqrt{\frac{16 \pi}{5}} \,,
\ee
where we set $x = (G M\omega/c^{3})^{2/3}$ and where $D$ is the luminosity distance to the observer. For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
\begin{subequations}\label{eq:SPAN}
\begin{align}
	A_{N}(f) &= \frac{G^{2}m^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,, \label{eq:ASPAN}\\
	\Psi_{\rm SPA}^{\rm N}(f) &= \phi_{0} - 2\pi f t_{0} - \frac{3}{128\nu v^{5}} \,, \label{eq:PsiSPAN} 
\end{align}
\end{subequations}
\SM{[Check sign of $\Psi$ ? Should have $<0$ 2nd der.] }
where we used the notation $v=(G m \pi f/c^{3})^{1/3}$ ($x=v^{2}$ with the SPA correspondence~\eqref{eq:deftfSPA}), and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\calM \equiv m\nu^{3/5}$, which is the only mass combination arising in the phase at the leading PN order.

@@

Additionally, we limit ourselves to a rigid model for the detector, namely we assume that the constellation remains in an equilateral configuration with fixed armlengths $L=2.5\times 10^{6}\mathrm{km}$, as proposed in~\cite{} \SM{[C]}. This simplified orbits neglect (iv) the effect of excentricity of the individual Keplerian orbits, which introduces corrections at the level $e^{2}$ (v) the effect of gravitational perturbations coming from other celestial bodies, such as the Earth, the quadrupole of the Sun, and the other planets.

@@

It is natural to define two transfer frequencies for the two relevant length scales for the delays, defined such that a wavelength fits within this length scale. Geometrical projection factors aside, these scales are the radius of the orbit around the Sun $R=1\text{au}=1.5\times 10^{8} \text{km}$ and the detector armlength $L=2.5\times 10^{6}\text{km}$ (in the configuration proposed in~\cite{} \SM{[C]}). This gives
\begin{subequations}\label{eq:transferfrequencies}
\begin{align}
	f_{R} &= 3.2\times10^{-4}\text{Hz} \,\\
	f_{L} &= 1.9\times 10^{-2}\text{Hz} \,.
\end{align}
\end{subequations}
These transfer frequencies will indicate where in the LISA band a delay term of the form $2\pi f d$ becomes larger than unity.\SM{[check if this appears elsewhere]}

@@

The error measure $\epsilon_{d}$ does not depend on signal-dependent timescales and can be directly read off the estimates above, with $\epsilon_{d} = \partial_{tf}G/(2\pi G)$. For the error measures $\epsilon_{\Psi}$ and $\epsilon_{A1}$, we can combine the above with the Newtonian timescales given in~\eqref{eq:timescalesN} for the inspiral phase of the signal. An important distinction needs to be made there, as the radiation reaction timescale $T_{f}$ grows rapidly towards lower frequencies as $f^{-11/6}$. If one considers systems that are arbitrarily far from merging, $T_{f}$ can be made arbitrarily large, leading the error measure $\epsilon_{\Psi}$ to take values arbitrarily larger than 1 where the perturbative treatment we adopted breaks down. This limit would describe systems observed in their deep inspiral and very slowly chirping, such as the population of galactic binaries that form one of the science targets of the LISA mission~\cite{} \SM{[C]}. Such signals would be very compact in Fourier domain, and are not covered by our formalism that assumes an extended Fourier spectrum. We will present in App.~\ref{} \SM{[R]} another version of our formalism based on discrete Fourier coefficients, that could apply better to this case \SM{[check: do we put this appendix or not ?]}. On the contrary, if one considers only systems that will merge during or shortly after the observation span of the mission, the starting frequency of the observations will be set by this limited lifetime of the source before merging and we can then bound $\epsilon_{\Psi}$ to be smaller than 1.

In the following we will consider only these merging binaries, which include the supermassive black hole mergers that we expect to detect~\cite{} \SM{[C]}. The population of stellar-origin binary black holes, proposed in~\cite{} \SM{[C]}, will have both sources rather far from merger and sources that will merge within a few years after the mission finishes. We will take a reference starting frequency corresponding to a time from merger $\Delta t = 10\text{yrs}$. Since the corrections turn out to be most important at the beginning of the signal, we will evaluate the timescales at this starting frequency. From the Newtonian relation~\eqref{eq:omegaphiN}, we have (using $\nu_{0} = 1/4$ for the equal-mass value): \SM{[check]}
\be
	f_{\rm start} = 2.9\times 10^{-5} \text{Hz} \left( \frac{\nu}{\nu_{0}} \right)^{-3/8} \left( \frac{m}{10^{6}M_{\odot}} \right)^{-5/8} \left( \frac{\Delta t}{10 \text{yr}} \right)^{-3/8} \,,
\ee
which gives in turn for the timescales
\begin{subequations}
\begin{align}
	T_{f} = 6.8\times10^{-2}\text{yr} \left( \frac{\nu}{\nu_{0}} \right)^{3/16} \left( \frac{m}{10^{6}M_{\odot}} \right)^{5/16} \left( \frac{\Delta t}{10 \text{yr}} \right)^{11/16} \,\\
	T_{A1} = 2.0\times10^{-4}\text{yr} \left( \frac{\nu}{\nu_{0}} \right)^{3/8} \left( \frac{m}{10^{6}M_{\odot}} \right)^{5/8} \left( \frac{\Delta t}{10 \text{yr}} \right)^{3/8} \,.
\end{align}
\end{subequations}
When inserting these expressions in~\eqref{eq:deffom}, in~\eqref{eq:estimatederivorb} and~\eqref{eq:estimatederivconst} we must also use $f_{\rm start}$ for the possible extra factors of $f$ or $f^{2}$ coming from the delay terms, which can lead to different overall scalings with mass ratio and total mass depending on where the signal starts in the band.

We summarize the results for the error estimates in Fig.~\ref{fig:fomLISA}. We show the magnitude of the figures of merit $\epsilon_{\Psi}$, $\epsilon_{A1}$, $\epsilon_{d}$ both for the orbital delay and the constellation response, for an equal-mass system with total masses of $m=10^{7} \Msol$, $10^{4} \Msol$ and $10^{2} \Msol$ and a starting frequency corresponding to $\Delta t =10 \text{yr}$ of observation. Since individual signals can show significant variation depending on the angular parameters, we show an average of $\ln \epsilon$ over the sky position, inclination and polarization, together with $\pm \sigma$ standard deviation, and overlay the estimates resulting from~\eqref{eq:estimatederivorb},~\eqref{eq:estimatederivconst} and~\eqref{eq:timescalesN}. Here and in the following, for completeness we show results on the frequency band from $10^{-5}\text{Hz}$ to $1Hz$, while the signals are expected to have little signal-to-noise ratio outside of the band from $10^{-4}\text{Hz}$ to $10^{-1}Hz$, which is sometimes taken as a reference~\cite{} \SM{[C]}.

@@

Having at hand generic estimates for where the higher-order corrections in the LISA response are expected to be relevant, we now demonstrate that including them indeed reduces the reconstruction error of the Fourier-domain waveform. We will use an example PhenomD~\cite{} \SM{[C]} waveform for an equal-mass and non-spinning system, and consider only two values for the total mass, $m=10^{7} \Msol$ and $m=10^{2} \Msol$. For this example, we will also focus on the orbital response and on a single-arm constellation response $y_{132}$, as TDI combinations will appear (within the approximations listed in Sec.~\ref{} \SM{[R]}) as linear combinations of these basic observables. On one hand, we perform a numerical inverse Fourier transform (IFFT) of the Fourier-domain PhenomD waveform, process the signal through the time-domain response~\eqref{eq:defresponseO} or~\eqref{eq:defresponsec}, and perform a numerical Fourier transform (FFT). This gives us the target Frourier-domain waveform, considered to be exact but for the possible artifacts of numerical Fourier transforms (most notably, oscillations induced by the necessary tapering of the signal). On the other hand, we process the Fourier-domain signal through the response summarized in Sec.~\ref{} \SM{[R]}, including various higher-order corrections. We then compare the output of the two procedures.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.48\linewidth]{plots/LISAtransferM1e7y12Lamp.pdf}
  \hspace{0.2cm}
  \includegraphics[width=.48\linewidth]{plots/LISAtransferM1e7y12Lphase.pdf}
  %
  \includegraphics[width=.48\linewidth]{plots/LISAerrorM1e7y12LcorrPsi.pdf}
  \hspace{0.2cm}
  \includegraphics[width=.48\linewidth]{plots/LISAerrorM1e7y12LcorrAd.pdf}
  \caption{Error in the transfer function for the basic observable $y_{12}$, for $M=10^{7} \Msol$.}
  \label{fig:LISAerrorM1e7const}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=.48\linewidth]{plots/LISAtransferM1e2y12Lamp.pdf}
  \hspace{0.2cm}
  \includegraphics[width=.48\linewidth]{plots/LISAtransferM1e2y12LReIm.pdf}
  %
  \includegraphics[width=.48\linewidth]{plots/LISAerrorM1e2y12LcorrPsi.pdf}
  \hspace{0.2cm}
  \includegraphics[width=.48\linewidth]{plots/LISAerrorM1e2y12LcorrAd.pdf}
  \caption{Transfer function and reconstruction error for the basic observable $y_{12}$, for $M=10^{2} \Msol$.}
  \label{fig:LISAerrorM1e2const}
\end{figure*}

@@

As proposed by several authors~\cite{} \SM{[C]}, if one performs the mode decomposition of the waveform~\eqref{eq:defmodes} not in a fixed inertial frame but rather in a time-dependent, rotating frame that follows the plane of the orbit, it is possible to restore much of the structure of a non-precessing waveform. In particular, one recovers qualitatively the hierarchy of mode amplitudes that prevails for non-precessing systems, with the modes $h_{22}$ and $h_{2,-2}$ being dominant.

We define $(\alpha, \beta, \gamma)$ are the Euler angles relating the precessing frame to the inertial frame in the $(z,y,z)$ convention. The $z$ axis of the inertial frame is chosen to be the direction of $\bm{J}$, the total angular momentum, which is almost constant\footnote{Except in the cases of transitional precession~\cite{Apostolatos+94}, where for high mass ratios and large antialigned spins the orbital angular momentum and the spin can almost cancel each other.}. The first two angles, $\alpha$ and $\beta$, are the two spherical angles tracking the direction of dominant radiation reaction, which during the inspiral follows essentially the normal to the orbital plane. The last angle $\gamma$ parametrizes the remaining freedom of rotation around this radiation axis. An important element of the construction of the precessing frame is the fixing of this third degree of freedom, for which we use the minimal rotation condition~\cite{} \SM{[C]}
\be\label{eq:gammadot}
	\dot{\gamma} = -\dot{\alpha}\cos \beta \,.
\ee
Quite generically, and in particular in the case of simple precession~\cite{Apostolatos+94, Kidder95} but also in the model we will take for the post-merger precession, the radiation axis precesses on a cone around the direction of $\bm{J}$, with $\alpha$ increasing, while $\beta$, the opening angle of the precession cone, is slowly varying.

@@

Several constructions for the precessing frame have been proposed~\cite{} \SM{[C]}. In the following, we will use a co-precessing frame based on the dominant eigenvector of the matrix representing the action of the angular momentum operator acting on the waveform modes, proposed in~\cite{} \SM{[C]} (see App.~\ref{app:wigner} for more details).

@@

When using the assumptions~\eqref{eq:approxsymmetryhlminusm}-\eqref{eq:approxzeronegativef}, one can derive a symmetry relation in the transfer functions themselves. From the explicit expression of the Wigner matrices, we have indeed (see App.~\ref{app:wigner})
\be
	\calD^{\ell *}_{-m,-m'} = (-1)^{m+m'}\calD^{\ell}_{mm'} \,.
\ee
Since for a function $g$ we have in general for its conjugate
\be
	\mathrm{FT}[g^{*}](f) = \tilde{g}(-f)^{*} \,,
\ee
we can write, using~\eqref{eq:approxsymmetryhlminusm},
\be
	\mathrm{FT}[\calD^{\ell *}_{mm'} h_{\ell m'}^{\rm P}](-f)^{*} = (-1)^{\ell+m+m'} \mathrm{FT}[\calD^{\ell *}_{-m,-m'} h_{\ell, -m'}^{\rm P}](f) \,,
\ee
or, for transfer functions,
\be
	\calT^{\ell}_{mm'}(f) = (-1)^{m+m'} \calT^{\ell *}_{-m,-m'}(-f) \,.
\ee
This means that such a model is required to cover only the positive frequency band and the values $m'>0$, the negative frequencies being deduced from
\begin{align}
	\tilde{h}^{\rm I}_{\ell m}(f) &= \sum_{m'>0} \calT^{\ell}_{mm'}(f) \tilde{h}^{\rm P}_{\ell, m'}(f) \text{ for } f>0 \,, \nn\\
	\tilde{h}^{\rm I}_{\ell m}(f) &= \sum_{m'>0} (-1)^{\ell+m+m'} \calT^{\ell *}_{-m,m'}(-f) \tilde{h}^{\rm P *}_{\ell, m'}(-f) \text{ for } f<0 \,.
\end{align}
When including only the leading harmonic $h^{\rm P}_{22}$ in the precessing-frame waveform (as is done for instance in PhenomP~\cite{} \SM{[C]}), we have for $f<0$
\begin{subequations}
\begin{align}
	\calT^{2}_{m,-2}(f) &= (-1)^{m} \calT^{\ell *}_{-m,2}(-f) \,, \\
	\tilde{h}^{\rm I}_{2 m}(f) &= (-1)^{m} \tilde{h}^{\rm I *}_{2,-m}(-f) \,. \label{eq:symmetryhIfor22only}
\end{align}
\end{subequations}
In general, when including a superposition of modes with different values of $m'$ in the model (for instance, the modes $h^{\rm P}_{22}$ and $h^{\rm P}_{21}$), this symmetry exists only at the level of the mode contributions, and there is no relation such as~\eqref{eq:symmetryhIfor22only}  at the level of $\tilde{h}^{\rm I}_{\ell m}$, as the complete mode reconstruction includes factors $(-1)^{\ell+m+m'}$ in the sum.

@@

The above decomposition~\eqref{eq:wignerrot}, and simplifying assumptions~\eqref{eq:approxsymmetryhlminusm}-\eqref{eq:approxzeronegativef} for the precessing-frame waveform, have been widely used for the existing models of waveforms from precessing binaries. For the inspiral phase of the SEOBNRv3 model~\cite{} \SM{[C]}, the frame is assumed to follow the instantaneous orbital plane and the waveform in this frame is represented by a spin-aligned expression, with spin-aligned components that are left to vary according to the precessing dynamics. The ringdown phase of the signal is treated differently, the waveform being first rotated back to an inertial frame and then matched to a superposition of quasi-normal modes in this final frame. The PhenomP model~\cite{} \SM{[C]} uses the decomposition~\eqref{eq:wignerrot} above (although directly in the Fourier domain, as we will detail below) with a precessing-frame waveform given by a spin-aligned PhenomD waveform~\cite{} \SM{[C]}. The SUA waveforms~\cite{} \SM{[C]}, focused on the inspiral phase, also use the decomposition~\eqref{eq:wignerrot} and a post-Newtonian non-precessing model for the modes $h_{\ell m}^{\rm P}$.

@@

In the first option, one applies directly the SPA, as described in Sec.~\ref{subsec:SPA}, to the product of the modulation and the signal. This is known~\cite{} \SM{[C]} to lead to possible pathologies, since the prefactor $1/\sqrt{\ddot{\phi}_{\rm prec}}$ can blow up due the precessing contributions to the phase, and is not used in waveform modelling.

>>>>

%\usepackage{showlabels}

@@

It was suggested by a number of authors~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} that the effect of the precession can be modeled, to a good approximation, by a time-dependent rotation of a non-precessing waveform. In this approach, one introduces a precessing frame following the evolution of the plane of the orbit, and approximates the waveform in this precessing frame by using a non-precessing model. Different prescriptions have been proposed for the construction of the precessing frame itself~\cite{} \SM{[C]}.

@@

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (at the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these Newtonian estimates will give useful orders of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $m=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/m^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}\label{eq:omegaphiN}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (Gm)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{G m \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
As for the leading-order time-domain amplitude of the $22$ mode, we have~\cite{BlanchetLiving}
\be\label{eq:a22N}
	a_{22}^{\rm N} (t) = \frac{2 G m \nu x}{D c^{2}} \sqrt{\frac{16 \pi}{5}} \,,
\ee
where we set $x = (G M\omega/c^{3})^{2/3}$ and where $D$ is the luminosity distance to the observer. For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
\begin{subequations}\label{eq:SPAN}
\begin{align}
	A_{N}(f) &= \frac{G^{2}m^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,, \label{eq:ASPAN}\\
	\Psi_{\rm SPA}^{\rm N}(f) &= \phi_{0} - 2\pi f t_{0} - \frac{3}{128\nu v^{5}} \,, \label{eq:PsiSPAN} 
\end{align}
\end{subequations}
\SM{[Check sign of $\Psi$ ? Should have $<0$ 2nd der.] }
where we used the notation $v=(G m \pi f/c^{3})^{1/3}$ ($x=v^{2}$ with the SPA correspondence~\eqref{eq:deftfSPA}), and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\calM \equiv m\nu^{3/5}$, which is the only mass combination arising in the phase at the leading PN order.

@@

Additionally, we limit ourselves to a rigid model for the detector, namely we assume that the constellation remains in an equilateral configuration with fixed armlengths $L=2.5\times 10^{6}\mathrm{km}$, as proposed in~\cite{LISA17}. This simplified orbits neglect (iv) the effect of excentricity of the individual Keplerian orbits, which introduces corrections at the level $e^{2}$ (v) the effect of gravitational perturbations coming from other celestial bodies, such as the Earth, the quadrupole of the Sun, and the other planets.

@@

It is natural to define two transfer frequencies for the two relevant length scales for the delays, defined such that a wavelength fits within this length scale. Geometrical projection factors aside, these scales are the radius of the orbit around the Sun $R=1\text{au}=1.5\times 10^{8} \text{km}$ and the detector armlength $L=2.5\times 10^{6}\text{km}$ (in the configuration proposed in~\cite{LISA17}). This gives
\begin{subequations}\label{eq:transferfrequencies}
\begin{align}
	f_{R} &= 3.2\times10^{-4}\text{Hz} \,\\
	f_{L} &= 1.9\times 10^{-2}\text{Hz} \,.
\end{align}
\end{subequations}
These transfer frequencies will indicate where in the LISA band a delay term of the form $2\pi f d$ becomes larger than unity.\SM{[check if this appears elsewhere]}

@@

The error measure $\epsilon_{d}$ does not depend on signal-dependent timescales and can be directly read off the estimates above, with $\epsilon_{d} = \partial_{tf}G/(2\pi G)$. For the error measures $\epsilon_{\Psi}$ and $\epsilon_{A1}$, we can combine the above with the Newtonian timescales given in~\eqref{eq:timescalesN} for the inspiral phase of the signal. An important distinction needs to be made there, as the radiation reaction timescale $T_{f}$ grows rapidly towards lower frequencies as $f^{-11/6}$. If one considers systems that are arbitrarily far from merging, $T_{f}$ can be made arbitrarily large, leading the error measure $\epsilon_{\Psi}$ to take values arbitrarily larger than 1 where the perturbative treatment we adopted breaks down. This limit would describe systems observed in their deep inspiral and very slowly chirping, such as the population of galactic binaries that form one of the science targets of the LISA mission~\cite{LISA17}. Such signals would be very compact in Fourier domain, and are not covered by our formalism that assumes an extended Fourier spectrum. We will present in App.~\ref{} \SM{[R]} another version of our formalism based on discrete Fourier coefficients, that could apply better to this case \SM{[check: do we put this appendix or not ?]}. On the contrary, if one considers only systems that will merge during or shortly after the observation span of the mission, the starting frequency of the observations will be set by this limited lifetime of the source before merging and we can then bound $\epsilon_{\Psi}$ to be smaller than 1.

In the following we will consider only these merging binaries, which include the supermassive black hole mergers that we expect to detect~\cite{LISA17}. The population of stellar-origin binary black holes, proposed in~\cite{Sesana16}, will have both sources rather far from merger and sources that will merge within a few years after the mission finishes. We will take a reference starting frequency corresponding to a time from merger $\Delta t = 10\text{yrs}$. Since the corrections turn out to be most important at the beginning of the signal, we will evaluate the timescales at this starting frequency. From the Newtonian relation~\eqref{eq:omegaphiN}, we have (using $\nu_{0} = 1/4$ for the equal-mass value): \SM{[check]}
\be
	f_{\rm start} = 2.9\times 10^{-5} \text{Hz} \left( \frac{\nu}{\nu_{0}} \right)^{-3/8} \left( \frac{m}{10^{6}M_{\odot}} \right)^{-5/8} \left( \frac{\Delta t}{10 \text{yr}} \right)^{-3/8} \,,
\ee
which gives in turn for the timescales
\begin{subequations}
\begin{align}
	T_{f} = 6.8\times10^{-2}\text{yr} \left( \frac{\nu}{\nu_{0}} \right)^{3/16} \left( \frac{m}{10^{6}M_{\odot}} \right)^{5/16} \left( \frac{\Delta t}{10 \text{yr}} \right)^{11/16} \,\\
	T_{A1} = 2.0\times10^{-4}\text{yr} \left( \frac{\nu}{\nu_{0}} \right)^{3/8} \left( \frac{m}{10^{6}M_{\odot}} \right)^{5/8} \left( \frac{\Delta t}{10 \text{yr}} \right)^{3/8} \,.
\end{align}
\end{subequations}
When inserting these expressions in~\eqref{eq:deffom}, in~\eqref{eq:estimatederivorb} and~\eqref{eq:estimatederivconst} we must also use $f_{\rm start}$ for the possible extra factors of $f$ or $f^{2}$ coming from the delay terms, which can lead to different overall scalings with mass ratio and total mass depending on where the signal starts in the band.

We summarize the results for the error estimates in Fig.~\ref{fig:fomLISA}. We show the magnitude of the figures of merit $\epsilon_{\Psi}$, $\epsilon_{A1}$, $\epsilon_{d}$ both for the orbital delay and the constellation response, for an equal-mass system with total masses of $m=10^{7} \Msol$, $10^{4} \Msol$ and $10^{2} \Msol$ and a starting frequency corresponding to $\Delta t =10 \text{yr}$ of observation. Since individual signals can show significant variation depending on the angular parameters, we show an average of $\ln \epsilon$ over the sky position, inclination and polarization, together with $\pm \sigma$ standard deviation, and overlay the estimates resulting from~\eqref{eq:estimatederivorb},~\eqref{eq:estimatederivconst} and~\eqref{eq:timescalesN}. Here and in the following, for completeness we show results on the frequency band from $10^{-5}\text{Hz}$ to $1Hz$, while the signals are expected to have little signal-to-noise ratio outside of the band from $10^{-4}\text{Hz}$ to $10^{-1}Hz$, which is sometimes taken as a reference~\cite{LISA17}.

@@

Having at hand generic estimates for where the higher-order corrections in the LISA response are expected to be relevant, we now demonstrate that including them indeed reduces the reconstruction error of the Fourier-domain waveform. We will use an example PhenomD~\cite{Khan+15,Husa+15} waveform for an equal-mass and non-spinning system, and consider only two values for the total mass, $m=10^{7} \Msol$ and $m=10^{2} \Msol$. For this example, we will also focus on the orbital response and on a single-arm constellation response $y_{132}$, as TDI combinations will appear (within the approximations listed in Sec.~\ref{} \SM{[R]}) as linear combinations of these basic observables. On one hand, we perform a numerical inverse Fourier transform (IFFT) of the Fourier-domain PhenomD waveform, process the signal through the time-domain response~\eqref{eq:defresponseO} or~\eqref{eq:defresponsec}, and perform a numerical Fourier transform (FFT). This gives us the target Frourier-domain waveform, considered to be exact but for the possible artifacts of numerical Fourier transforms (most notably, oscillations induced by the necessary tapering of the signal). On the other hand, we process the Fourier-domain signal through the response summarized in Sec.~\ref{} \SM{[R]}, including various higher-order corrections. We then compare the output of the two procedures.

@@

\begin{figure}
  \centering
  \includegraphics[width=.98\linewidth]{plots/LISAtransferM1e7y12LReIm.pdf}
  %
  \includegraphics[width=.98\linewidth]{plots/LISAerrorM1e7y12L.pdf}
  \caption{Error in the transfer function for the basic observable $y_{12}$, for $M=10^{7} \Msol$.}
  \label{fig:LISAerrorM1e7const}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.98\linewidth]{plots/LISAtransferM1e2y12LReIm.pdf}
  %
  \includegraphics[width=.98\linewidth]{plots/LISAerrorM1e2y12L.pdf}
  \caption{Transfer function and reconstruction error for the basic observable $y_{12}$, for $M=10^{2} \Msol$.}
  \label{fig:LISAerrorM1e2const}
\end{figure}

@@

As proposed by several authors~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11}, if one performs the mode decomposition of the waveform~\eqref{eq:defmodes} not in a fixed inertial frame but rather in a time-dependent, rotating frame that follows the plane of the orbit, it is possible to restore much of the structure of a non-precessing waveform. In particular, one recovers qualitatively the hierarchy of mode amplitudes that prevails for non-precessing systems, with the modes $h_{22}$ and $h_{2,-2}$ being dominant.

We define $(\alpha, \beta, \gamma)$ are the Euler angles relating the precessing frame to the inertial frame in the $(z,y,z)$ convention. The $z$ axis of the inertial frame is chosen to be the direction of $\bm{J}$, the total angular momentum, which is almost constant\footnote{Except in the cases of transitional precession~\cite{Apostolatos+94}, where for high mass ratios and large antialigned spins the orbital angular momentum and the spin can almost cancel each other.}. The first two angles, $\alpha$ and $\beta$, are the two spherical angles tracking the direction of dominant radiation reaction, which during the inspiral follows essentially the normal to the orbital plane. The last angle $\gamma$ parametrizes the remaining freedom of rotation around this radiation axis. An important element of the construction of the precessing frame is the fixing of this third degree of freedom, for which we use the minimal rotation condition~\cite{Boyle+11}
\be\label{eq:gammadot}
	\dot{\gamma} = -\dot{\alpha}\cos \beta \,.
\ee
Quite generically, and in particular in the case of simple precession~\cite{Apostolatos+94, Kidder95} but also in the model we will take for the post-merger precession, the radiation axis precesses on a cone around the direction of $\bm{J}$, with $\alpha$ increasing, while $\beta$, the opening angle of the precession cone, is slowly varying.

@@

Several constructions for the precessing frame have been proposed~\cite{} \SM{[C]}. In the following, we will use a co-precessing frame based on the dominant eigenvector of the matrix representing the action of the angular momentum operator acting on the waveform modes, proposed in~\cite{OShaughnessy+11} (see App.~\ref{app:wigner} for more details \SM{[R]}).

@@

When using the assumptions~\eqref{eq:approxsymmetryhlminusm}-\eqref{eq:approxzeronegativef}, one can derive a symmetry relation in the transfer functions themselves. From the explicit expression of the Wigner matrices, we have indeed (see App.~\ref{app:wigner})
\be
	\calD^{\ell *}_{-m,-m'} = (-1)^{m+m'}\calD^{\ell}_{mm'} \,.
\ee
Since for a function $g$ we have in general for its conjugate
\be
	\mathrm{FT}[g^{*}](f) = \tilde{g}(-f)^{*} \,,
\ee
we can write, using~\eqref{eq:approxsymmetryhlminusm},
\be
	\mathrm{FT}[\calD^{\ell *}_{mm'} h_{\ell m'}^{\rm P}](-f)^{*} = (-1)^{\ell+m+m'} \mathrm{FT}[\calD^{\ell *}_{-m,-m'} h_{\ell, -m'}^{\rm P}](f) \,,
\ee
or, for transfer functions,
\be
	\calT^{\ell}_{mm'}(f) = (-1)^{m+m'} \calT^{\ell *}_{-m,-m'}(-f) \,.
\ee
This means that such a model is required to cover only the positive frequency band and the values $m'>0$, the negative frequencies being deduced from
\begin{align}
	\tilde{h}^{\rm I}_{\ell m}(f) &= \sum_{m'>0} \calT^{\ell}_{mm'}(f) \tilde{h}^{\rm P}_{\ell, m'}(f) \text{ for } f>0 \,, \nn\\
	\tilde{h}^{\rm I}_{\ell m}(f) &= \sum_{m'>0} (-1)^{\ell+m+m'} \calT^{\ell *}_{-m,m'}(-f) \tilde{h}^{\rm P *}_{\ell, m'}(-f) \text{ for } f<0 \,.
\end{align}
When including only the leading harmonic $h^{\rm P}_{22}$ in the precessing-frame waveform (as is done for instance in PhenomP~\cite{Hannam+13}), we have for $f<0$
\begin{subequations}
\begin{align}
	\calT^{2}_{m,-2}(f) &= (-1)^{m} \calT^{\ell *}_{-m,2}(-f) \,, \\
	\tilde{h}^{\rm I}_{2 m}(f) &= (-1)^{m} \tilde{h}^{\rm I *}_{2,-m}(-f) \,. \label{eq:symmetryhIfor22only}
\end{align}
\end{subequations}
In general, when including a superposition of modes with different values of $m'$ in the model (for instance, the modes $h^{\rm P}_{22}$ and $h^{\rm P}_{21}$), this symmetry exists only at the level of the mode contributions, and there is no relation such as~\eqref{eq:symmetryhIfor22only}  at the level of $\tilde{h}^{\rm I}_{\ell m}$, as the complete mode reconstruction includes factors $(-1)^{\ell+m+m'}$ in the sum.

@@

The above decomposition~\eqref{eq:wignerrot}, and simplifying assumptions~\eqref{eq:approxsymmetryhlminusm}-\eqref{eq:approxzeronegativef} for the precessing-frame waveform, have been widely used for the existing models of waveforms from precessing binaries. For the inspiral phase of the SEOBNRv3 model~\cite{Pan+13,BTB16}, the frame is assumed to follow the instantaneous orbital plane and the waveform in this frame is represented by a spin-aligned expression, with spin-aligned components that are left to vary according to the precessing dynamics. The ringdown phase of the signal is treated differently, the waveform being first rotated back to an inertial frame and then matched to a superposition of quasi-normal modes in this final frame. The PhenomP model~\cite{Hannam+13} uses the decomposition~\eqref{eq:wignerrot} above (although directly in the Fourier domain, as we will detail below) with a precessing-frame waveform given by a spin-aligned PhenomD waveform~\cite{Khan+15,Husa+15}. The SUA waveforms~\cite{KCY14,Chatziioannou+16,Chatziioannou+17}, focused on the inspiral phase, also use the decomposition~\eqref{eq:wignerrot} and a post-Newtonian non-precessing model for the modes $h_{\ell m}^{\rm P}$.

@@

In the first option, one applies directly the SPA, as described in Sec.~\ref{subsec:SPA}, to the product of the modulation and the signal. This is known (see e.g.~\cite{KCY13}) to lead to possible pathologies, since the prefactor $1/\sqrt{\ddot{\phi}_{\rm prec}}$ can blow up due the precessing contributions to the phase, and is not used in waveform modelling.
